Towards an Automated Sportscasting System for Chess

### 2011-05-18 paper/paper.tex
Added segmented data

\begin{itemize}

\item Different applications of similar techniques
\item Differences from sportscasting for, e.g., RoboCup
\item Text about approach
\end{itemize}

\section{Related Work}
\paragraph{Deep learning.} These techniques attempt to capture complex relationships between input variables in order to transform the input into a different representation which better expresses the relevant structures that make up the input. Specifically, they compose many layers of non-linearities in order to compactly represent complex functions which can model these relationships. They have been used successfully for many different types of tasks \cite{ng-mnist,ng-vision,ng-nlp}.

@@

\paragraph{Automatic chess commentary.}

\section{Data Sources}

We obtained chess games, both annotated and unannotated, from various websites in the PGN format \cite{pgn}, the universal format for compactly representing chess games, because we anticipated that many open source PGN parsers would be available.

The following websites contain chess games with annotations: \url{http://wwwu.uni-klu.ac.at/gossimit/c/tactic.htm} \url{http://webplaza.pt.lu/public/ckaber/Chess.htm#Chess training}. We obtained example games from them to feed into our classifier for the supervised component of our system.

For the unsupervised component, we obtained as many chess games as possible from a variety of websites like chessgames.com \cite{other sites}. In all, we obtained over five million untagged games spread over 600,000 files and 5 GB of data. On top of that, we obtained over 5000 tagged games.

\subsection{Data Sanitation}	
In order to ensure consistency and integrity over the games obtained from disparate sources, we sanitized them in multiple ways. We parsed each game in the data set and removed information that would confused the parser. For example, some games had English or German comments by humans that were not structured and thus meaningful for machine learning. Others had alternate branches, sometimes up to multiple levels (alternate branches within alternate branches). Still others represented games in a non-standard way (For example, pawn promotions were represented without the “=”), which we had to fix. By fixing our input games, we ensured a consistent format for our machine learning system.

In addition, because the games were from multiple sources, there was significant overlap in games (approximately 50%). Therefore, we also went through all of the downloaded games and eliminated duplicate games: Multiple games that had the same initial position and move sequence were considered redundant and only one of each were kept.

\subsection{Parsing}
We tried various open source PGN parsers, each succumbing to flaws like mishandling uncommon cases (like promotions and special moves like en passant). Finally we chose Chesspresso parser, which we fixed and modified for our needs. Chesspresso takes a PGN file as input and returns a data structure that represents the initial position of the board and the subsequent moves (and all metadata necessary to recreate the PGN file). We also extended it to be more robust to variations in input format, in addition to being able to export PGNs in a canonical form, which was crucial for deduplication.

\section{Features}

\subsection{Sparse Autoencoding}
\subsection{Manual baseline}
As a first step towards classification, we selected chess-specific manual features from a given chess position-previousmove pair.  The position-based features involve information such as where the pieces are on the board and potential moves, while the move-based features involve binary information such as if the move is a capturing move or a check.  We list all our hand-selected manual features in Table REF.

@@

@@

\begin{table*}
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Name} & \textbf{Description} \\ \hline
\tt{attackPossible} & The current player has opportunity to attack. \\
\tt{backrankWeakness} & There exists a future checkmate by a rook or queen on the edges of the board. \\
\tt{bishopPair} & There exists a decisive strategic advantage for the player with two bishops. \\
\tt{clearance} & There exists a maneuver to break the pawn structure. \\ 
\tt{closure} & There exists a maneuver to fix the pawn structure. \\
\tt{decoy} & There exists a maneuver to distract a crucial piece. \\
\tt{defensePressure} & There exists a maneuver to break the opposing King’s defense. \\
\tt{deflection} & There exists a maneuver to deflect an attack.  \\ 
\tt{discoveredAttack} & There exists a maneuver to allow for a discovered attack (an attack that was blocked by a friendly piece.) \\
\tt{doubleAttack} & There exists a move that simultaneously attacks at least two crucial squares. \\
\tt{drawCombinationsStalemate} & There exists a combination that allows for decisive material advantage or immediate checkmate. \\
\tt{enclosedKings} & There exists a maneuver to trap the opposing King. \\
\tt{freeingBlockingSquares} & There exists a move to free a crucial square. \\
\tt{goodEndgameChances} & There exists a move to allow for a favorable endgame. \\
\tt{inBetweenMove} & There exists an interrupting move that allows for decisive material advantage of immediate checkmate. \\ 
\tt{interference} & There exists a maneuver to block a crucial piece. \\ 
\tt{interruption} & There exists a maneuver that interrupts the opponent’s tactics. \\
\tt{kingInCenter} & There exists a decisive strategic advantage with the player with the King in the center of the board. \\
\tt{kingToBadSquares} & There exists a combination that forces the opposing King to move to a bad square. \\
\tt{longDiagonals} & There exists a strategic advantage to the player that controls squares on the long diagonal. \\
\tt{matingAttack} & There exists a decisive mating attack. \\
\tt{openingClosingDiagonals} & There exists a maneuver to clear a diagonal. \\
\tt{openingClosingFiles} & There exists a maneuver to clear a file. \\
\tt{overload} & There exists a combination that makes a defending piece defend too many squares at once. \\
\tt{passedPawn} & There exists a combination that allows for a passed pawn (a pawn with no opposing pawns in front of it) \\
\tt{pawnPromotion} & There exists a combination that allows for a pawn promotion. \\
\tt{perpetualCheck} & There exists a combination that forces a perpetual check/draw. \\
\tt{pin} & There exists a combination that allows for a pin. \\
\tt{pullingPiecesToBadSquares} & There exists a combination that forces an opposing piece to move to a bad square. \\
\tt{queenCapture} & There exists a combination to capture the opposing Queen. \\
\tt{quietMates} & There exists a quiet move (one that looks inconsequential on the surface) leading to mate. \\
\tt{removalOfDefence} & There exists a maneuver to remove an opponent’s defender. \\
\tt{unpinning} & There exists a maneuver to unpin a piece. \\
\tt{wrongBishop} & The opposing player’s only bishop is limited in use because of the pawn structure of the game. \\ \hline
\end{tabular} 
\caption{List of high-level themes we attempt to automatically identify.}
\end{table*}

@@

\section{Experiments}
\subsection{Results}
\subsection{Discussion}
We discovered that there is an inherent trade off between how general the approach is and the quality of the outcomes. In pursuing this project, we attempted to maintain an approach that is as general as possible so that it can be applied to other contexts. However, in attempting to maintain generality, we sacrificed access to powerful but chess-specific tools like game engines, resulting in a classifier that is more surface-level (yet slower) than existing chess analysis tools that leverage the game engine.

\section{Future Work}
There are three areas in which future work can make progress on automated sportscasting. It could improve the performance of the autoencoder so that it can be trained on a larger data set. Furthermore, it could focus on trying different methods to improve the classification quality by leveraging chess-specific tools and features like chess engines and evaluation tools. Moreover, it could investigate an orthogonal path by applying our general approach to other games, including more complex ones like StarCraft and simpler ones like Checkers.

@@

>>>>

\section{Related Work}
\begin{itemize}
\item Different applications of similar techniques
\item Differences from sportscasting for, e.g., RoboCup
\item Text about approach
\end{itemize}

\paragraph{Deep learning.} These techniques attempt to capture complex relationships between input variables in order to transform the input into a different representation which better expresses the relevant structures that make up the input. Specifically, they compose many layers of non-linearities in order to compactly represent complex functions which can model these relationships. They have been used successfully for many different types of tasks \cite{ng-mnist,ng-vision,ng-nlp}.

@@

\paragraph{Automatic chess commentary.} There exist techniques to automatically annotate chess games which are heavily dependent on rule-based logic and use of a chess-engine to determine which parts of a specific chess game are salient~\cite{cambridge-chess-annotation}. As a departure from the rule-based approach to automatic chess commentary, Guid et al. (\cite{learning-positional-features}) discuss the use of argument based machine learning, which analyzes reasoning where arguments for and against a certain claim are produced and evaluated (in the context of chess, why a certain move was played over others.) Their system, much like the one presented in this paper, identifies complex positional features of a given chess game.

However, we identified two shortcomings of the AMBL system.  First, it relies on experts to provide arguments as to why a certain position exemplifies (or doesn’t) a particular positional feature.  While this is a step up from defining a classification rule by hand, it is infeasible to iterate this procedure multiple times.  Moreover, Guid described that

\section{Data Sources and Processing}
We obtained both annotated and unannotated records of chess games from various web sites in the PGN format \cite{pgn}, a widely-used format for compactly representing chess games. A large number of games as well as software and tools that use the PGN format are freely available.

The already-annotated games come from databases of short snippets of chess games,\footnote{\url{http://wwwu.uni-klu.ac.at/gossimit/c/tactic.htm}} \footnote{\url{http://webplaza.pt.lu/public/ckaber/Chess.htm#Chess training}} intended for chess players to learn from by example, which are organized by particular themes that the students are to identify. These provided a useful repository of both interesting high-level concepts we could identify (we provide a listing in Table~\ref{table:themes}), as well as boards and moves corresponding to each one.

We also looked for as many chess games as possible from a variety of web sites, such as \url{chessgames.com}. We easily obtained over five million untagged games spread over 600,000 files and 5 GB of data. The number of annotated game snippets we obtained was slightly over 5,000.

\subsection{Data Pre-Processing}	
In order to ensure consistency and integrity over the games obtained from disparate sources, we sanitized them in multiple ways. We parsed each game in the data set and removed information that would confuse the parser. For example, some games had English or German comments by humans which we did not plan to use, and sometimes resulted in parsing errors due to specific characters present in the comments. Others had alternate branches, sometimes up to multiple levels (alternate branches within alternate branches), which the parser could not handle properly. Still others represented games in a non-standard way (for example, pawn promotions were represented without the ``=''), which we had to fixed.

Since we obtained the games from many sources, there was significant overlap in games (approximately 50% of the unlabeled games collected were redundant). Therefore, we also went through all of the downloaded games and eliminated duplicate games: multiple games that had the same initial position and move sequence were considered redundant and only one of each were kept.

\subsection{Parsing}
We tried various open source PGN parsers, each succumbing to flaws like mishandling uncommon cases (like promotions and special moves like en passant). We based our system on Chesspresso,\footnote{url{http://www.chesspresso.org/}} a Java library for writing chess programs, which we fixed and modified for our needs. Chesspresso takes a PGN file as input and returns a data structure that represents the initial position of the board and the subsequent moves (as well as all metadata necessary to recreate the PGN file). We also extended it to be more robust to variations in input format, in addition to being able to export PGNs in a canonical form, which was crucial for removing duplicates.

\section{Features}
\subsection{Manually Generated Features}
As a first step towards classification, we selected chess-specific manual features from a given position-move pair.  The position-based features involve information such as where the pieces are on the board and potential moves, while the move-based features involve binary information such as if the move is a capturing move or a check.  We list all our hand-selected manual features in Table REF.

@@

\subsection{Sparse Autoencoding}

@@

\begin{table*}
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Name} & \textbf{Description} \\ \hline
\tt{attackPossible} & The current player has opportunity to attack. \\
\tt{backrankWeakness} & There exists a future checkmate by a rook or queen on the edges of the board. \\
\tt{bishopPair} & There exists a decisive strategic advantage for the player with two bishops. \\
\tt{clearance} & There exists a maneuver to break the pawn structure. \\ 
\tt{closure} & There exists a maneuver to fix the pawn structure. \\
\tt{decoy} & There exists a maneuver to distract a crucial piece. \\
\tt{defensePressure} & There exists a maneuver to break the opposing King’s defense. \\
\tt{deflection} & There exists a maneuver to deflect an attack.  \\ 
\tt{discoveredAttack} & There exists a maneuver to allow for a discovered attack (an attack that was blocked by a friendly piece.) \\
\tt{doubleAttack} & There exists a move that simultaneously attacks at least two crucial squares. \\
\tt{drawCombinationsStalemate} & There exists a combination that allows for decisive material advantage or immediate checkmate. \\
\tt{enclosedKings} & There exists a maneuver to trap the opposing King. \\
\tt{freeingBlockingSquares} & There exists a move to free a crucial square. \\
\tt{goodEndgameChances} & There exists a move to allow for a favorable endgame. \\
\tt{inBetweenMove} & There exists an interrupting move that allows for decisive material advantage of immediate checkmate. \\ 
\tt{interference} & There exists a maneuver to block a crucial piece. \\ 
\tt{interruption} & There exists a maneuver that interrupts the opponent’s tactics. \\
\tt{kingInCenter} & There exists a decisive strategic advantage with the player with the King in the center of the board. \\
\tt{kingToBadSquares} & There exists a combination that forces the opposing King to move to a bad square. \\
\tt{longDiagonals} & There exists a strategic advantage to the player that controls squares on the long diagonal. \\
\tt{matingAttack} & There exists a decisive mating attack. \\
\tt{openingClosingDiagonals} & There exists a maneuver to clear a diagonal. \\
\tt{openingClosingFiles} & There exists a maneuver to clear a file. \\
\tt{overload} & There exists a combination that makes a defending piece defend too many squares at once. \\
\tt{passedPawn} & There exists a combination that allows for a passed pawn (a pawn with no opposing pawns in front of it) \\
\tt{pawnPromotion} & There exists a combination that allows for a pawn promotion. \\
\tt{perpetualCheck} & There exists a combination that forces a perpetual check/draw. \\
\tt{pin} & There exists a combination that allows for a pin. \\
\tt{pullingPiecesToBadSquares} & There exists a combination that forces an opposing piece to move to a bad square. \\
\tt{queenCapture} & There exists a combination to capture the opposing Queen. \\
\tt{quietMates} & There exists a quiet move (one that looks inconsequential on the surface) leading to mate. \\
\tt{removalOfDefence} & There exists a maneuver to remove an opponent’s defender. \\
\tt{unpinning} & There exists a maneuver to unpin a piece. \\
\tt{wrongBishop} & The opposing player’s only bishop is limited in use because of the pawn structure of the game. \\ \hline
\end{tabular}
\caption{List of high-level themes we attempt to automatically identify, along with explanations}
\label{table:themes}
\end{table*}

@@

\section{Experiments}
\subsection{SVM Based Structured Classifier}

In our experiments, we processed the input data into multimaps from game prefixes to tags. We then segmented the data into three disjoint sets: 90% for training, 5% as the validation set (for tuning our classifiers), and 5% as the test set. The segmenting was done by randomly sampling from the set of all labeled games.

Afterwards, we constructed lists of positive and negative labels (mappings from tags to games that have that tag, and games that do not have that tag). We then built SVM models using the training data. In testing, we fed that to the SVM to generate, for each test game and tag, a boolean prediction of whether that tag applies to that game. As a result, we constructed a SVM-based multiple-class classifier that attempts to predict the tags that characterize an input game.

We measured the accuracy of our model using various measures like false positive rate, false negative rate, precision, recall, and F1. In addition, we compared the predictions made by SVM to those made randomly where each game was assigned a tag at random based on the proportion of the games that have that tag. The random predictions were made 1,000 times, with the predictions made independently at random, and the error rates were averaged.

\subsection{Results}
Since each game has very few tags, most of the binary classifications are negative. As a result, the model is heavily biased toward negative results. Our results bear this disposition:

INSERT TABLE HERE

The basic classifier generates classifications with relatively high precision compared to a random classifier. However, the recall is much lower than that of the random classifier. As a result, the F1 score (a combination of the precision and recall scores) of the SVM-based model is no better than that of the random model.

We then extended the tagging to also incorporate information about the player

\subsection{Discussion}
We discovered that there is an inherent trade off between how general the approach is and the quality of the outcomes. In pursuing this project, we attempted to maintain an approach that is as general as possible so that it can be applied to other contexts. However, in attempting to maintain generality, we sacrificed access to powerful but chess-specific tools like game engines, resulting in a classifier that is more surface-level (yet slower) than existing chess analysis tools that leverage the game engine.

\section{Future Work}
There are three areas in which future work can make progress on automated sportscasting. It could improve the performance of the autoencoder so that it can be trained on a larger data set. Furthermore, it could focus on trying different methods to improve the classification quality by leveraging chess-specific tools and features like chess engines and evaluation tools. Moreover, it could investigate an orthogonal path by applying our general approach to other games, including complex ones like StarCraft and simpler ones like Checkers.

@@

stack of work:
Segmented:
True positive: 13 (0.001252)
False positive: 2 (0.000193)
True negative: 10213 (0.983438)
False negative: 157 (0.015118)
Precision/Recall: 0.866667, 0.076471
F1 Score: 0.140541
True positive: 12841 (0.001236)
False positive: 154930 (0.014919)
True negative: 10060070 (0.968712)
False negative: 157159 (0.015133)
Precision/Recall: 0.076539, 0.075535
F1 Score: 0.076034

Unsegmented:
True positive: 11 (0.002087)
False positive: 0 (0.000000)
True negative: 5106 (0.968880)
False negative: 153 (0.029032)
Precision/Recall: 1.000000, 0.067073
F1 Score: 0.125714
True positive: 22387 (0.004248)
False positive: 145688 (0.027645)
True negative: 4960312 (0.941236)
False negative: 141613 (0.026872)
Precision/Recall: 0.133196, 0.136506
F1 Score: 0.134831

\begin{table*}
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{True Positive} & \textbf{False Positive} & \textbf{True Negative} & \textbf{False Negative} & \textbf{Precision} & \textbf{Recall} & \textbf{F1} \\ \hline
Position features & Piece location, potential moves per piece, material count, square control. \\
Move features & Capturing move, recapturing move, en passant move, long castling move, short castling move, check move, stalemate move, checkmate move, promotion move.
\end{tabular} 
\caption{List of chess-specific manual features per position.}
\end{table*}
